From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-28 paper.tex
Rewritten discussion of with-me-ness results

Algorithm~\ref{alg:with-me-ness} provides a formal way of computing the level of
with-me-ness $\mathcal{W}$ between two time points $[t_{start}, t_{end}]$. A
notable difference with the original definition by Sharma~\etal is that, at a
given time $t$, the task $task(t)$ performed by the robot may elicit more than
one attentional target; thus, at a given time, more than one location can be
regarded as possible \emph{expected} focuses of attention for the human. For
example, a robot which is writing, could typically elicit gazes to its hand as
well as to its head. A human looking at either of these locations would be
considered to be \emph{with} the robot in terms of
interaction\footnote{Considering a probabilistic model of
attention expectations would be an interesting extension of this metric.}. Also notable, we exclude from the computation of
$\mathcal{W}$ all of the periods of time where the user's focus of attention can
not be estimated (typically because the user's
face is not visible at those times).

\subsection{Experimental Interpretation}

Over the course of the experiment presented at section~\ref{sec:expe}, the robot
controller would associate a set of expected attentional targets to the task being
performed. For instance, while the robot was waiting for the child's handwriting
demonstration, the expected attentional target was the tablet (since the child
was supposed to write there), and while the robot was writing itself, the
attentional targets were expected to be either on the tablet or on the robot.
These expected targets, forming altogether the robot's attentional {\it a
priori}, are depicted on Figure~\ref{fig:with-me-ness} as the green lines. They
are used to compute the with-me-ness. With-me-ness can be calculated over the
whole interaction (Table~\ref{tab:results-with-me-ness}) or over shorter time
windows (bottom graph on Figure~\ref{fig:with-me-ness}). Shorter time windows
are interesting for two purposes: to analyse the level of with-me-ness in
relation to specific interaction episodes; to allow a measurement of
with-me-ness by the robot \emph{over the course} of the interaction
(\emph{in-the-moment} measurement) -- in this later case, one may typically want
to consider a sliding time window.

@@

Figure~\ref{fig:with-me-ness} can be further analyzed:\fixme{This paragraph
needs to be adapted}
The matching curve (cyan color) shows the overlapping between the
expected and the real focuses of attention. As can be noticed there is a
fluctuation at the beginning of the interaction as well as just before the
activity switching. These chaotic gaze manifestation is what we call an
automatic discovery of events, when the user experiences an uncertainty in the
present situation.

The first one of the previous two situations presented above is due to the
novelty effect the subjects experience for the first time in front of the system
and according to the rest of participants suggests that the adaptation time is
around the first 2 minutes of interaction. In the second case, an unexpected
robot behavior produces such reaction due to an activity switching, the story
telling.

It is important to consider that in some occasions during the interaction, none
of the focuses of attention were within the field of view due to the time
transition from one focus to another resulting in an ambiguity. In order to deal
with these few cases it is assumed that the minimum distance between the height
of the field of view (cone) and the targets corresponds to the most likely focus
of attention. However, we also want to distinguish such cases when the subject
is directly not looking where the defined targets are located, but towards a
region far from the interaction field. These cases are not common but they can
be easily identified by setting a horizontal threshold to the field of view.

A relevant trend that can be observed is the continuous `look for approval'
situation where the subjects look at the experimenter in order to provide
feedback about the robot's response before providing their own one (see yellow
boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
becomes more frequent in several cases, for instance when the experimenter makes
a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
to a more complicated word.

Once the turn-taking is established, a pattern becomes visible during the
subject's feedback, when an attempt is performed: Such pattern shows a high
intermittent gaze frequency from the tablet to the selection. It also reveals
that the subject is using the demonstration model in the selection tablet to
provide a better answer to the robot. In fact, subjects decrease the response
time after each shift as well as the gaze frequency towards the model suggesting
an assimilation over time. Moreover, when a new demonstration is selected for
playing, the same behavior is replicated (see yellow boxes in figure
\ref{fig:realExpected}).

If we consider the case shown in section \ref{sec:fromFaceTo}, evaluating the
subject's progression together with the time on task may be an indicator of
engagement: If the subject answers so quickly with respect to the previous
attempts with a poorer handwriting may be the case she is not engaged anymore.
In such hypothetical case an activity switch can be easily implemented in a
robot architecture. It confirms the findings reported in \cite{anzalone} where
time response to event is proposed to account quality of interaction. 
%
%However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the
%demonstration provided can reveal that the challenge is too big or no gaze, the
%exactly the opposite. In this case, the robot response can be adapted to the
%situation. Similar, but least relevant results are obtained using the
%correctness (figure \ref{fig:errorR}) of the child's feedback after each
%attempt. 
%
%\begin{figure}
%  \centering
%  \begin{tabular}{@{}c@{}}
%    \includegraphics[width=.7\linewidth,height=100pt]{gazeR} \\
%    \small (a) Occurrences of a gaze shift.
%  \end{tabular}\label{fig:gazeR}
%
%  \vspace{\floatsep}
%
%  \begin{tabular}{@{}c@{}}
%    \includegraphics[width=.7\linewidth,height=100pt]{errorR} \\
%    \small (b) Word correctness.
%  \end{tabular}\label{fig:errorR}
%
%  \caption{ \small Linear regression of all subject attempts against the gaze
%  shift tablet-model (a) and the correctness as the average euclidean distance
%  of each letter in its corresponding eigenspace with respect to a reference
%  letter (b). The gray region defines the confidence interval.}
%  
%\end{figure}
%
%
%This information can be used to estimate the child's state during the
%interaction. For instance, if we consider that a low level of engagement is
%produced by extremely high or low demanding situations, it is possible to assess
%the engagement level during the interaction based on gaze shift and word
%correctness.

It is also important to notice that the initiative of the robot into the
learning process in a turn-taking activity produces an effect in the engagement
of the human partner. However, experience suggests that a robot that learns
evoke a positive action in the partner. But the opposite, a robot that is not
able to progress over time, produces the contrary effect.

@@

Compared to engagement, one of the strength of with-me-ness is its specificity:
it is well-defined, we can formalize it, and
as such, it is valuable to assess and compare how users are willing or able to
interact with a robot. Besides, we have hopefully demonstrated in this article
that with-me-ness is an operational \emph{in-the-moment} metric that can be also
used as feed-back to the robot controller to build richer, more adaptive
interactive behaviors for our robots.

>>>>

Algorithm~\ref{alg:with-me-ness} provides a formal way of computing the level of
with-me-ness $\mathcal{W}$ between two time points $[t_{start}, t_{end}]$. A
notable difference with the original definition by Sharma~\etal is that, at a
given time $t$, the task $task(t)$ performed by the robot may elicit more than
one attentional target; thus, at a given time, more than one location can be
regarded as possible \emph{expected} focuses of attention for the human. For
example, a robot which is writing, could typically elicit gazes to its hand as
well as to its head. A human looking at either of these locations would be
considered to be \emph{with} the robot in terms of
interaction\footnote{Considering a probabilistic model of
attention expectations (an attention distribution) would be an interesting
extension of this metric.}. Also notable, we exclude from the computation of
$\mathcal{W}$ all of the periods of time where the user's focus of attention can
not be estimated (typically because the user's face is not visible at those
times).

\subsection{Experimental Measure \& Interpretation}

\begin{table}[t]
    \centering
    \caption{Mapping between the interaction phases and the expected
        attentional targets.}

    \begin{tabular}{ll}
        \toprule
        {\bf Phase} & {\bf Expected targets} \\
        \midrule
        Presentation & {\sf robot} \\ 
        Waiting for word to write & {\sf secondary tablet} \\ 
        Writing word & {\sf tablet}, {\sf robot} \\ 
        Waiting for feedback & {\sf tablet}, {\sf secondary tablet} \\ 
        Story telling & {\sf robot} \\ 
        Bye & {\sf robot} \\ 
        \bottomrule
    \end{tabular}
    \label{tab:targets}
\end{table}

Over the course of the experiment presented at section~\ref{sec:expe}, the robot
controller would associate a set of expected attentional targets to the phase of
the interaction (Table~\ref{tab:targets}). For instance, while the robot was
waiting for the child's handwriting demonstration (``\textit{Waiting for
feedback}''), the expected attentional target of the child was the tablet (since
the child was supposed to write there) or the secondary tablet (that displayed a
template of the word, used as reference by the child). These expected targets
form altogether the robot's attentional {\it a priori} and are used to compute
the with-me-ness.  They are represented over time by the green lines on
Figure~\ref{fig:with-me-ness}. With-me-ness can be calculated over the whole
interaction (see corresponding results in Table~\ref{tab:results-with-me-ness})
or over shorter time windows. Shorter time windows are interesting for two
purposes: to analyse the level of with-me-ness in relation to specific
interaction episodes; to allow a measurement of with-me-ness by the robot
\emph{over the course} of the interaction (\emph{in-the-moment} measurement) --
in this later case, one may typically want to consider a sliding time window.

@@

The with-me-ness plotted at the bottom of Figure~\ref{fig:with-me-ness} is in
fact computed on a sliding window of 30 seconds, and thus gives a picture of
``how well the child is following the robot's expectations'' at that time. As
seen, the with-me-ness computed at run-time by the robot (blue line) is
generally lower than the ground-truth (orange line, based on video-annotations),
and sometimes quite off, such as during episode marked ``A'': during that phase,
one can notice that the attention is mostly directed to undefined target {\sf
Other}, likely a consequence of inaccurate head detection.  This kind of error
(inaccurate head pose estimation) is the main source of discrepancy between the
ground-truth and the attention distribution measured by the robot: ignoring all
the episodes where the child's gaze is measured to be directed to {\sf Other},
we indeed obtain levels of with-me-ness close to the ground-truth (over the six
subjects, $M=87.5, SD=4.6$).

A chart like Figure~\ref{fig:with-me-ness} remains an useful tool to analyse the
interaction, and several observations can be made out of it: as said, the green
lines represent how the robot imagine, at a given time, the attention
distribution of the child. They also provide an accurate picture of the overall
turn-taking as viewed by the robot: for instance, the episode ``B'' on
Figure~\ref{fig:with-me-ness} corresponds to one of the ``\emph{Robot writing}''
episodes, surrounded by ``\emph{Waiting for feedback}'' phases like ``C''; episode
``D'' corresponds to the story telling; etc. In terms of interaction, the large
variance of the duration of these phases reflects the fact that this child would
sometimes take a lot of time to send feedback to the robot, and sometimes, on
the contrary, be very quick.

Looking at the ground-truth focus of attention (orange lines), the first
striking observation is that this child did generally \emph{closely} follow what
the robot was expecting: in that regard, it seems that the child was very much
engaged into the interaction (we discuss in the next section the exact
relationships between with-me-ness and engagement). The only major exception is
the story-telling phase (episode ``D''): the child was seemingly not captivated by the
first half of the story, and his attention was not directed towards the robot.

Another interesting observation pertains to the facilitator: as one can see,
this child only rarely turned to the facilitator, indicating that the
interaction and the task were meaningful and easy for him to follow alone.

More subtle patterns and events can also be observed: for instance, during the
feedback phases like episode ``C'', we
can notice numerous gaze shifts between the tablet (where the child write) and the
secondary tablet (that showed a template of the word). The episode ``B'' (robot
writing) is also interesting: the child did not look at the robot, and instead
remained focused on the secondary tablet. This situation is typically useful for
the robot to detect as it may want to adapt its behavior to recover the child's
attention.

%A relevant trend that can be observed is the continuous `look for approval'
%situation where the subjects look at the experimenter in order to provide
%feedback about the robot's response before providing their own one (see yellow
%boxes in figure \ref{fig:realExpected}). Such transition tablet-experimenter
%becomes more frequent in several cases, for instance when the experimenter makes
%a suggestion (orange box in figure \ref{fig:realExpected}), in this case, switch
%to a more complicated word.
%
%Once the turn-taking is established, a pattern becomes visible during the
%subject's feedback, when an attempt is performed: Such pattern shows a high
%intermittent gaze frequency from the tablet to the selection. It also reveals
%that the subject is using the demonstration model in the selection tablet to
%provide a better answer to the robot. In fact, subjects decrease the response
%time after each shift as well as the gaze frequency towards the model suggesting
%an assimilation over time. Moreover, when a new demonstration is selected for
%playing, the same behavior is replicated (see yellow boxes in figure
%\ref{fig:realExpected}).

%If we consider the case shown in section \ref{sec:fromFaceTo}, evaluating the
%subject's progression together with the time on task may be an indicator of
%engagement: If the subject answers so quickly with respect to the previous
%attempts with a poorer handwriting may be the case she is not engaged anymore.
%In such hypothetical case an activity switch can be easily implemented in a
%robot architecture. It confirms the findings reported in \cite{anzalone} where
%time response to event is proposed to account quality of interaction. 
%
%However, a high amount of gaze shift (figure \ref{fig:gazeR}) towards the
%demonstration provided can reveal that the challenge is too big or no gaze, the
%exactly the opposite. In this case, the robot response can be adapted to the
%situation. Similar, but least relevant results are obtained using the
%correctness (figure \ref{fig:errorR}) of the child's feedback after each
%attempt. 
%
%\begin{figure}
%  \centering
%  \begin{tabular}{@{}c@{}}
%    \includegraphics[width=.7\linewidth,height=100pt]{gazeR} \\
%    \small (a) Occurrences of a gaze shift.
%  \end{tabular}\label{fig:gazeR}
%
%  \vspace{\floatsep}
%
%  \begin{tabular}{@{}c@{}}
%    \includegraphics[width=.7\linewidth,height=100pt]{errorR} \\
%    \small (b) Word correctness.
%  \end{tabular}\label{fig:errorR}
%
%  \caption{ \small Linear regression of all subject attempts against the gaze
%  shift tablet-model (a) and the correctness as the average euclidean distance
%  of each letter in its corresponding eigenspace with respect to a reference
%  letter (b). The gray region defines the confidence interval.}
%  
%\end{figure}
%
%
%This information can be used to estimate the child's state during the
%interaction. For instance, if we consider that a low level of engagement is
%produced by extremely high or low demanding situations, it is possible to assess
%the engagement level during the interaction based on gaze shift and word
%correctness.

%It is also important to notice that the initiative of the robot into the
%learning process in a turn-taking activity produces an effect in the engagement
%of the human partner. However, experience suggests that a robot that learns
%evoke a positive action in the partner. But the opposite, a robot that is not
%able to progress over time, produces the contrary effect.

@@

Compared to engagement, one of the strength of with-me-ness is its specificity:
it is well-defined (the only important variable -- that needs to be reported --
is the mapping {\it phase/expected attentional target}, \ie our
Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
assess and compare how users are willing or able to interact with a robot.
Besides, we have hopefully demonstrated in this article that with-me-ness is an
operational \emph{in-the-moment} metric that can be also used as feedback to the
robot controller to build richer, more adaptive interactive behaviors for our
robots.
